---
title: "Moose Abundance and Density in the Oil Sands Region of Alberta"
author: "Marcus Becker"
date: "March 14, 2019"
output: 
  html_document:
    highlight: zenburn
    theme: readable
    code_folding: hide
    self_contained: false
---

```{r include=TRUE, echo=TRUE, message=FALSE, warning=FALSE, eval=TRUE, cache=FALSE}

# Load packages
library(sf)
library(tidyverse)
library(leaflet)
library(htmlwidgets)
library(leaflet.extras)
library(splitstackshape)
library(rmapshaper)
# if needed:
# devtools::install_github("r-spatial/leafgl")
library(leafgl)
library(Hmisc) 
library(colourvalues)

# Import data
ab_wmu <- st_read("./SpatialData/AB_WMU-OSR-Intersect.shp", 
                  stringsAsFactors = FALSE, quiet = TRUE)

pixel_1km <- st_read("./SpatialData/ABMI_Grid_1KM_OSM_v3.shp",
                     stringsAsFactors = FALSE, quiet = TRUE)

ab_osr <- st_read("./SpatialData/AB_WMU-OSR-Dissolve.shp",
                  stringsAsFactors = FALSE, quiet = TRUE)

gov_obs <- read_csv("./Data/Raw/OSM_Ungulate_Survey_Data_2013-2018.csv")
abmi_pred <- read_csv("./Data/Raw/Moose_km2-summary.csv")
gov_dens <- read_csv("./Data/Raw/OSM_Ungulate_Survey_Data_2013-2018_Density.csv")

# Prepare aerial observation data
gov_obs_sf <- gov_obs %>%
  # Change NAs to WMU 515
  mutate(WMU = if_else(is.na(WMU), 515, WMU)) %>%
  filter(SPECIES == "MOOSE -MOOS") %>%
  select(-DATE_1) %>%
  mutate(WMU = as.character(WMU)) %>%
  # Convert to sf object
  st_as_sf(coords = c("LONG", "LAT"), crs = 4326, agr = "constant") %>%
  expandRows("COUNT") %>%
  # Jitter for better visualization of hotspots w/ multiple animals sighted
  st_jitter(0.01)

# Prepare WMU polygons
ab_wmu <- ab_wmu %>%
  separate(WMUNIT_COD, into = c("zeroes", "WMU"), sep = 2, remove = FALSE) %>%
  select(-zeroes) %>%
  filter(WMU %in% gov_obs_sf$WMU) %>%
  mutate(WMU = as.numeric(WMU)) %>%
  # Join density estimates
  left_join(gov_dens, by = "WMU") %>%
  # Set projection
  st_transform("+init=epsg:4326") 
  # ms_simplify(sys = TRUE)

# Prepare ABMI 1km pixel polygons
pixel_1km_aoi <- pixel_1km %>%
  separate(WMUNIT_COD, into = c("zeroes", "WMU"), sep = 2, remove = FALSE) %>%
  select(-zeroes) %>%
  filter(WMU %in% gov_obs_sf$WMU) %>%
  select(GRID_LABEL, WMUNIT_NAM, WMU) %>%
  rename(LinkID = GRID_LABEL) %>%
  # Join in Moose prediction data
  left_join(abmi_pred, by = "LinkID") %>%
  # Set projection
  st_transform("+init=epsg:4326") %>%
  st_cast("POLYGON") %>%
  ms_simplify(sys = TRUE) %>%
  # Subset
  filter(!is.na(Curr)) %>%
  mutate(q_curr = as.numeric(cut2(Curr, g = 100))) %>%
  select(q_curr)

# Project and clean ab_osr
ab_osr <- ab_osr %>% 
  st_transform("+init=epsg:4326") %>%
  ms_simplify(sys = TRUE)

knitr::opts_chunk$set(fig.width = 12, fig.height = 10)

```

The first map is a simple chloropleth map of estimated moose density in each 
WMU within the Oil Sands Monitoring Region. Density is estimated using 
observation data from aerial surveys. The WMUs within the region are surveyed 
on a rotating basis, with approximately five surveyed each year. 

Click on each WMU in the map to display the name, code, and most recent
estimated moose density. 

```{r include=TRUE, echo=TRUE, message=FALSE, warning=FALSE, eval=TRUE}

# Define colours
pal_1 <- colorNumeric(palette = "YlOrBr", domain = ab_wmu$Density)

# Generate leaflet map

map_1 <-
ab_osr %>%
  leaflet() %>%
  addTiles() %>%
  addProviderTiles("Stamen.TerrainBackground") %>%
  addProviderTiles("Esri.WorldImagery", group = "Imagery") %>%
  addFullscreenControl() %>%
  addResetMapButton() %>%
  addScaleBar(position = "bottomright", 
              options = scaleBarOptions(imperial = FALSE)) %>%
  setView(lng = -113.07, lat = 56.53, zoom = 7) %>%
  
  # Polygon layers
  addPolygons(color = "#070707", weight = 1, smoothFactor =  0.2, 
              opacity = 2.0) %>%
  
  addPolygons(data = ab_wmu, color = "#070707", weight = 0.8,
              smoothFactor = 0.5, opacity = 1.0, fillOpacity = 0.75, 
              fillColor = ~ pal_1(Density), 
              popup = paste("WMU Name:", 
                            "<b>", ab_wmu$WMUNIT_NAM, "</b>", "<br>",
                            "WMU Code:", 
                            "<b>", ab_wmu$WMU, "</b>", "<br>",
                            "Moose Density:", 
                            "<b>", ab_wmu$Density, "</b>", "per km2", "<br>", 
                            "<br>",
                            "<i>", "Last surveyed in", 
                            ab_wmu$Survey_Year, "<i>"),
              group = "Moose Density - Aerial Surveys") %>%
  
  # Legend layer
  addLegend(data = ab_wmu, position = "bottomleft", pal = pal_1, bins = 6,
            values = ~Density, 
            title = "Moose Density:<br>Animals per km2",
            opacity = 0.9, group = "Moose Density - Aerial Surveys") %>%
  
  # Layers control
  addLayersControl(overlayGroups = c("Imagery",
                                     "Moose Density - Aerial Surveys"),
                   options = layersControlOptions(collapsed = FALSE)) %>%
  
  hideGroup("Imagery")

map_1

```

The second map displays the ABMI's predicted relative abundance of moose in the 
Oil Sands Monitoring Region using camera trap data and species-specific habitat
modeling. Abundance is predicted at the 1km grid cell size. 

The user can also toggle between the predicted abundance layer and the raw 
aerial survey observation data (presented as a heatmap). 


```{r include=TRUE, echo=TRUE, message=FALSE, warning=FALSE, eval=TRUE}

# Define colours
c1 <- c("blue", "white", "red")
c2 <- colorRampPalette(c1)
colours <- c2(100)[pixel_1km_aoi$q_curr]

colours1 <- (convert_colors(colours)) / 255

pal_2 <- colorNumeric(c("blue", "white", "red"),
                      domain = pixel_1km_aoi$q_curr)

# Generate map

map_2 <- 
  ab_wmu %>%
  leaflet() %>%
  addTiles() %>%
  addProviderTiles("Stamen.TerrainBackground") %>%
  addProviderTiles("Esri.WorldImagery", group = "Imagery") %>%
  addFullscreenControl() %>%
  addResetMapButton() %>%
  addScaleBar(position = "bottomright", 
              options = scaleBarOptions(imperial = FALSE)) %>%
  # addMapPane(name = "WMU Boundaries", zIndex = 420) %>%
  # addMapPane(name = "Everything Else", zIndex = 410) %>%
  setView(lng = -113.07, lat = 56.53, zoom = 7) %>%
  
  # Polygon layers
  addPolygons(color = "#070707", weight = 2,
              smoothFactor = 0.2, opacity = 2.0,
              popup = paste("WMU Name:", 
                            "<b>", ab_wmu$WMUNIT_NAM, "</b>", "<br>",
                            "WMU Code:", 
                            "<b>", ab_wmu$WMU, "</b>", "<br>",
                            "Moose Density:", 
                            "<b>", ab_wmu$Density, "</b>", "per km2", "<br>", 
                            "<br>",
                            "<i>", "Last surveyed in", 
                            ab_wmu$Survey_Year, "<i>")) %>%
  
  addGlPolygons(data = pixel_1km_aoi,
                color = colours1,
                group = "ABMI Relative Abundance Predictions",
                opacity = 0.75) %>%
  
  addHeatmap(data = gov_obs_sf, radius = 10, blur = 18, minOpacity = 4.0,
             group = "Aerial Observations Heatmap") %>%
  
  # Legend layer
  addLegend(data = pixel_1km_aoi, position = "bottomleft", pal = pal_2, 
            bins = 10, values = ~q_curr, 
            title = "Relative Abundance of Moose", 
            opacity = 0.9, group = "ABMI Relative Abundance Predictions") %>%
  
  # Layers control
  addLayersControl(overlayGroups = "Imagery",
                   baseGroups = c("ABMI Relative Abundance Predictions",
                                  "Aerial Observations Heatmap"),
                   options = layersControlOptions(collapsed = FALSE)) %>%
  
  hideGroup(c("Imagery", 
              "Aerial Observations Heatmap"))
  
map_2
  
```































